---
title: "Class 11"
author: "Vanesa Fernandez"
format: pdf
---

Today, before delving into structure prediction with AlfaFold we will finish off our previous lab10 "comparative structure analysis" section.

```{r}
installed.packages("bio3d")

library(bio3d)
id <- "1ake_A"

aa <- get.seq(id)
print(aa)

```


```{r}
b <- blast.pdb(aa)
```


```{r}
attributes(b)
```

```{r}
head(b$hit.tbl)
```

```{r}
hits <- plot(b)
```

```{r}
attributes(hits)
```


These are the top hits that we like from our blast results:
```{r}
hits$pdb.id
```


```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)

```

I downloaded all ADK structures in the PDB database but viewing them is difficult as they need to be aligned and superimpossed

To install BiocManager package from CRAN
```{r}

##install.packages("BiocManager")

##BiocManager::install() ## <- use it to install any bioconductor package, BiocManager::install.
```


```{r}
# Align related PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```


```{r}
pdbs
```


## Principal component analysis 

```{r}
# Perform Principal Component Analysis on the aligned PDB structures
pc <- pca(pdbs)
plot(pc)
```

```{r}
# Plot the first two principal components (PC1 vs PC2)
plot(pc, pc.axes = c(1,2))
```

## To examine in more detail what PC1 (or any PC) is capturing here we can plot the loadings or make a wee movie (trajectory) of moving along PC1
#note: the zero is the average... the more you move in any direction you do away from the avergae towards whatever the PC is attributed to.

```{r}
mktrj(pc, pc=1, file="pc1.pdb")
```

##Visualize the molecule structure in mo* https://molstar.org/viewer/
#1) Open this file in mol*
#2) Start animation 
#3) Export animation if need be

##This is the first PC, and doing it this way is much easier than doing it one by one 
