---
title: "Class 09 Halloween Mini-Project"
author: "Vanesa Fernandez A59026769"
format: gfm
---

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

>Q3. What is your favorate candy?, what is it’s `winpercent` value?

```{r}
candy["Milky Way", ]$winpercent
```

```{r}
candy["charleston chew",]
```

```{r}
candy["Charleston Chew",]$winpercent
candy["Dums Dums",]$winpercent
```


```{r}
library(dplyr)

candy |>
  filter(rownames(candy) %in% c("Dums Dums", "Twix")) |>
  select(winpercent)
```

>Q4. What is the winpercent value for “Kit Kat”?
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy |>
  filter(rownames(candy) %in% c("Kit Kat",
                                "Tootsie Roll Snack Bars")) |>
  select(winpercent)
```
The `%in%` operator is useful for checking the intersection of two vectors. 
```{r}
c("barry", "liz", "chandra") %in% c("paul", 
                                   "alice", "liz")
```
```{r}
candy |>
  filter (winpercent >75) |>
  filter(pricepercent < 0.5)
```

```{r}
##install.packages("skimr")
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
>Q7. What do you think a zero and one represent for the candy$chocolate column?
> Q8. Plot a histogram of winpercent values

We can do this a few different ways: with base R `hist()` or with `ggplot()`. Let's try both...

```{r}
hist(candy$winpercent, breaks=10)
```
> Q9. Is the distribution of winpercent values symmetrical?

```{r}
library (ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10) +
  theme_bw()
```

>Q10. Is the center of the distribution above or below 50%? below

```{r}
summary(candy$winpercent)
```
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
inds <- as.logical(candy$chocolate)
candy [inds,]$winpercent
```

```{r}
inds <- candy$chocolate == 1
candy[inds,]$winpercent
```
## alternatively : 
```{r}
candy |>
  filter(chocolate == 1) |>
  select(winpercent)

```

```{r}
inds <- candy$chocolate == 1
choc.win <- candy[inds,]$winpercent

inds <- candy$fruity == 1
fruit.win <- candy[inds,]$winpercent
```

Then I could compare these
```{r}
summary(fruit.win)
summary(choc.win)
```


```{r}
t.test(fruit.win, choc.win)
```


>Q13. What are the five least liked candy types in this set?

There are two related function that are useful here `sort()` and `order()`
```{r}
play <- c(2,1, 5,3)
sort(play)
```

```{r}
order(play) ## I dont understand this part
```
```{r}
play [order(play)]
```

```{r}
l <- c("c", "a", "b")
sort(l)
```

```{r}
order(l)
```
```{r}
n <- c("d", "a")
n[order(n)]
```

```{r}
inds <- order(candy$winpercent)
head(candy[inds,])
```

```{r}
head(candy[order(candy$winpercent),], n=5)

```
OR: 
```{r}
candy %>% arrange(winpercent) %>% head(5)
```
Which approach do you prefer and why? `candy %>% arrange(winpercent) %>% head(5)` because It's easier to remember and shorter

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
inds <- order(candy$winpercent, decreasing = T)
head(candy[inds,], 5)
```
```{r}
ggplot(candy) +
  aes(y=rownames(candy), x= winpercent) +
  geom_col()
```


>Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
ggplot(candy) +
  aes(x= winpercent, y=rownames(candy)) +
  geom_col()
```


```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent, y=reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
ggplot(candy) +
  aes(x=winpercent, y=reorder(rownames(candy), winpercent), fill=chocolate) +
  geom_col()
```
I want to define my own colors
I am start with a place-holder vector of colors all "black"

```{r}
mycols <- rep("black", nrow(candy))
mycols [as.logical(candy$chocolate)] <- "chocolate"
mycols [as.logical(candy$bar)] <- "brown"
mycols [as.logical(candy$fruity)] <- "pink"

##Favorite candy and color:
mycols[rownames(candy) == "Milky Way"] <- "yellow"

mycols
```

> Q. Color your favorite candy with your favorite color

```{r}
ggplot(candy) +
  aes(x=winpercent, y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```
>Q17. What is the worst ranked chocolate candy?
Sixlets
>Q18. What is the best ranked fruity candy?
Starburst

Plot winpercent vs pricepercent
```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text(col=mycols)
```
```{r}
##install.packages(ggrepel)

library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
A: chocolate and fruty 

>Q23. Similarly, what two variables are most positively correlated?
A: bar and chocolate 

## Exploring the correlation structure
```{r}
cij <- cor(candy)
##install.packages("corrplot")
library(corrplot)
corrplot (cij)
```

## Principal Component Analysis 
```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col=mycols, pch=16)
```

```{r}
pca$rotation[,1]
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```
```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)

p
```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
fruity, hard and pluribus. It makes sense, it indicates people don't like these ones. 


