---
title: "Class 12"
author: "Vanesa Fernandez"
format: pdf
---

## Population Scale Analysis 
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3 expression.

>> Exploring the data 

```{r}
## reading the data 

expr <- read.table ("rs8067378_ENSG00000172057.6.txt")

head(expr)

```
> Q. How many samples do we have?

```{r}
##How many total samples?
nrow(expr)
```

> Q. How many of each genotype?

```{r}
sample_size <- table(expr$geno)
print(sample_size)
```
## Making a figure to communicate the results

```{r}
library(ggplot2)
```

Visualizing the data -- Let's make a boxplot where we compare genotype vs  expression levels  
```{r}
##helps to open the data table in another window to retrieve the name of the columns for x, y axes in the plot
ggplot(expr) + 
  aes(x=geno, y= exp, fill=geno) + ## coloring filling by genotype
  geom_boxplot(notch=TRUE) ## returns a list that contains a variety of statistics, including the median values. This list can be saved into an object and then accessed for getting the stats I need.

```
So far, with this plain plot and without running stats we can see that G/G genotype is associated with a reduced expression of ORMDL3 gene. 

To extract the medians from the boxplot created above, use the boxplot function separately to save it as an object:

```{r}
# Create a boxplot and save the output to an object
boxplot_stats <- boxplot(exp ~ geno, data = expr, plot = FALSE) ## creates the boxplot but does not plot it. Instead, it returns a list of statistics 
```


```{r}
# Extract the median values - 3rd row
medians <- boxplot_stats$stats[3, ] ## boxplot_stats$stats is a matrix where each row corresponds to a statistical value (e.g., the minimum, lower quartile, median, upper quartile, and maximum). The median is always in the second row.

print(medians)
```

```{r}
# Combine medians and sample sizes into a data frame
results <- data.frame(
  Genotype = names(sample_size), # Genotype names from the sample_size table
  SampleSize = as.numeric(sample_size), # Sample sizes converted to numeric
  MedianExpression = medians
)

print(results)
```

Alternatively ...
```{r}
#Extract medians using the `median()` function 
median_values <- tapply(expr$exp, expr$geno, median)

print(median_values)
```


